\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}

\newcommand{\bbF}{\mathbb{F}}
\newcommand{\bbQ}{\mathbb{Q}}
\newcommand{\bbZ}{\mathbb{Z}}
\newcommand{\cO}{\mathcal{O}}
\newcommand{\rhob}{\bar{\rho}}
\newcommand{\Gal}{\mathrm{Gal}}
\newcommand{\Frob}{\mathrm{Frob}}
\newcommand{\Jac}{\mathrm{Jac}}
\newcommand{\Tr}{\mathrm{Tr}}
\newcommand{\Nm}{\mathrm{Nm}}
\newcommand{\End}{\mathrm{End}}
\newcommand{\chicyc}{\chi_{11}}
\DeclareMathOperator{\GSp}{GSp}

\title{A mod-11 modularity match over $\bbQ(\sqrt5)$ \\ from a $(1,11)$-polarized abelian surface}
\author{(working note)}
\date{June 2026 (corrected 29 June 2026)}

\begin{document}
\maketitle

\begin{abstract}
We exhibit computational evidence for Serre's modularity conjecture over the real quadratic
field $K=\bbQ(\sqrt5)$. Starting from a point of the Gross--Popescu moduli of
$(1,11)$-polarized abelian surfaces, we construct a genus-2 curve $C$ over $K$ whose Jacobian
$A=\Jac(C)$ is geometrically simple ($\End_{\bar K}A=\bbZ$) with good reduction at both primes
above $11$. The mod-$11$ representation $A[11]$ splits over $K$ as a sum of two irreducible
$2$-dimensional pieces $\rho_1,\rho_2$ with cyclotomic determinant --- a splitting forced not
by extra endomorphisms but by the polarization, since $11$ is the polarization degree. Working
entirely mod $11$ inside a space of Hilbert modular forms computed via a definite quaternion
algebra, we analyze whether $\rho_1$ and $\rho_2$ are realized by Hilbert modular eigenforms
of parallel weight $2$.

\medskip\noindent\textbf{Correction (2026-06-29).} An earlier version of this note claimed a
match (modularity of $\rho_1,\rho_2$ up to a quadratic twist at $2$, at level
$\mathfrak p\cdot(2)$). \emph{That claim was a false positive and is retracted.} The
acceptance criterion used --- nonvanishing of a ``twist-robust'' intersection $V$ --- is too
weak, and the correct per-eigenform test fails (Section~\ref{sec:refutation}). The structural
results on $A$ and on $A[11]=\rho_1\oplus\rho_2$ (Sections~\ref{sec:surface}--\ref{sec:rep})
are unaffected; the modularity of $\rho_1,\rho_2$ over $\bbQ(\sqrt5)$ is, at present,
\emph{not} established by this computation. What we \emph{can} do is derive, by elimination,
the precise modularity statement that must hold (weight, character, level up to one bounded
2-adic exponent); see Section~\ref{sec:derived}.
\end{abstract}

\section{Introduction}

Serre's modularity conjecture, a theorem over $\bbQ$ (Khare--Wintenberger), remains largely
open over totally real fields. In its Hilbert-modular form it predicts that every odd,
irreducible, continuous representation
$\rhob\colon \Gal(\bar K/K)\to \mathrm{GL}_2(\bar\bbF_\ell)$ of the absolute Galois group of a
totally real field $K$ arises from a Hilbert modular eigenform. This note records a concrete
instance over $K=\bbQ(\sqrt5)$ with $\ell=11$, obtained by the following pipeline:
\[
\text{$(1,11)$ Gross--Popescu point} \;\longrightarrow\; \text{abelian surface }A/K
\;\longrightarrow\; A[11]=\rho_1\oplus\rho_2 \;\longrightarrow\; \text{Hilbert modular form}.
\]
The point of the construction is that the abelian surface comes with a canonical mod-$11$
structure (the $(1,11)$-polarization), which is exactly what makes $A[11]$ reducible into
$2$-dimensional pieces and supplies interesting $\rhob$'s to test.

\section{The abelian surface}\label{sec:surface}

Let $K=\bbQ(\sqrt5)$, $\cO_K=\bbZ[\tfrac{1+\sqrt5}2]$, of class number $1$ and discriminant
$5$. The prime $11$ splits in $K$.

The Gross--Popescu construction (\texttt{arXiv:math/9902017}) realizes a $(1,11)$-polarized
abelian surface in $\mathbb P^{10}$ from a point of the sextic ``odd $2$-torsion'' locus
$D_2\subset\mathbb P^4$. Inverting the associated theta map numerically yields a period matrix
$\tau$, hence (via its principally polarized partner) a genus-2 curve $C/K$. We work with the
curve {\bf Curve A}, with absolute Igusa--Siegel invariants (normalized $J_2=1$)
\[
\Big[\,1,\ \tfrac{243125\sqrt5-482787}{1459240},\
\tfrac{-54798934\sqrt5+127710391}{2229718720},\
\tfrac{182422196780\sqrt5-406668692089}{8517525510400},\
\tfrac{38134182372761\sqrt5-85270624049895}{65073894899456000}\,\Big].
\]

\begin{proposition}\label{prop:A}
$A=\Jac(C)$ satisfies:
\begin{enumerate}
\item[(i)] $A$ is geometrically simple with $\End_{\bar K}(A)=\bbZ$ (no real multiplication);
\item[(ii)] $A$ has good reduction at both primes of $K$ above $11$;
\item[(iii)] away from $11$, the conductor of $A$ has odd part $\mathfrak p$, a prime of $K$
of norm $66179$, at which $A$ has multiplicative reduction (conductor exponent $1$);
\item[(iv)] $A$ also has (model-)bad reduction at the inert prime $(2)$.
\end{enumerate}
\end{proposition}

These are verified by computation: (i) via the geometric endomorphism algebra (CHIMP), (ii)
and (iii) via the Igusa valuation criterion and conductor exponents.

\section{The mod-11 representation}\label{sec:rep}

Let $\rhob\colon\Gal(\bar K/K)\to\GSp_4(\bbF_{11})$ be the representation on $A[11]$, and let
$\chicyc$ denote the mod-$11$ cyclotomic character. We summarize the structure, all of which
is verified over the $290$ good primes $P$ of $K$ with $N(P)\le 2000$ by reducing $C$ and
factoring the Frobenius characteristic polynomial $h_P(x)\in\bbF_{11}[x]$.

\begin{proposition}\label{prop:struct}
\begin{enumerate}
\item[(a)] For every such $P$, $h_P$ factors over $\bbF_{11}$ as a product of two quadratics:
its factor type is always $[1,1,2]$ or $[1,1,1,1]$, and never $[2,2]$, $[1,3]$, or $[4]$.
\item[(b)] $\rhob$ has no $1$-dimensional $K$-subquotient: there is no Galois character
$\lambda$ that is, at every prime, a root of the ``linear part'' of $h_P$ (an exhaustive search
over Hecke characters of $K$ of order dividing $10$ and every plausible conductor gives at best
$90/132$ agreement).
\item[(c)] Consequently $A[11]$ is, up to semisimplification, a sum
$\rho_1\oplus\rho_2$ of two $2$-dimensional representations, each with
$\det\rho_i=\chicyc$ (hence odd). The splitting is supplied by the polarization: the kernel
$K(\mathcal L)\cong(\bbZ/11)^2$ of the $(1,11)$-polarization is a $K$-rational $2$-dimensional
subspace of $A[11]$. The prime $11$ is special precisely because it is the polarization degree;
at other primes $\ell$ the image of $\Jac(C)[\ell]$ is the full $\GSp_4(\bbF_\ell)$.
\item[(d)] $\rho_1,\rho_2$ are irreducible over $K$, are not isomorphic, and are not quadratic
twists of one another; their image is large (not dihedral/CM: the surface trace vanishes at
only $\approx 10\%$ of primes).
\end{enumerate}
\end{proposition}

Because $\det\rho_i=\chicyc$, the two traces are the roots of an explicit quadratic. Writing
$h_P(x)=x^4+Ax^3+Bx^2+Cx+D$ (so $D=N(P)^2$ and $C=N(P)\,A$), one has
\begin{equation}\label{eq:fingerprint}
\{\,\mathrm{tr}\,\rho_1(\Frob_P),\ \mathrm{tr}\,\rho_2(\Frob_P)\,\}
=\text{roots of }\ y^2 + A\,y + \big(B-2N(P)\big)\ \ (\mathrm{mod}\ 11).
\end{equation}
We call $\big(\Tr_P,\Nm_P\big):=\big(-A,\ B-2N(P)\big)\bmod 11$ the {\it fingerprint} of
$\{\rho_1,\rho_2\}$ at $P$. A sample:
\[
\begin{array}{c|cccccccc}
N(P) & 49 & 169 & 29 & 29 & 31 & 31 & 41 & 41\\\hline
\Tr_P & 9 & 6 & 7 & 9 & 5 & 0 & 9 & 2\\
\Nm_P & 9 & 5 & 1 & 7 & 4 & 10 & 0 & 3
\end{array}
\]

\section{The modularity match}

By good reduction at $11$ (Proposition~\ref{prop:A}(ii)) each $\rho_i$ is finite-flat at $11$,
so the predicted Serre weight is parallel weight $2$, and the predicted level is prime to $11$
and divides the conductor of $A$, i.e.\ $\mathfrak p\cdot (2)^{a}$ for some $a\ge0$.

Let $S_2(\mathfrak n)$ be the space of Hilbert cusp forms over $K$ of parallel weight $2$ and
level $\mathfrak n$, computed via the definite quaternion algebra ramified at the two
archimedean places, with its commuting Hecke operators $T_P$ (Brandt matrices), which are
integral. Reduce everything mod $11$. If a Hilbert newform $f$ has
$\rhob_{f,\lambda}\cong\rho_i$ for some degree-one prime $\lambda\mid 11$ of its Hecke field,
then its reduction is an eigenvector $v$ in $S_2(\mathfrak n)\otimes\bbF_{11}$ with
$T_P v = t_P\,v$ where $t_P$ is a root of \eqref{eq:fingerprint}; that is,
$\big(T_P^2 - \Tr_P\,T_P + \Nm_P\big)v=0$.

Empirically the matching forms realize $\rho_i$ only after a {\it quadratic twist}
$\varepsilon$ ramified at $2$, so $t_P$ is replaced by $\varepsilon(P)\,t_P=\pm t_P$. We
therefore intersect the twist-robust kernels: set
\begin{equation}\label{eq:V}
V \;=\; \bigcap_{P}\ \Big[\ \ker\!\big(T_P^2 - \Tr_P\,T_P + \Nm_P\big)\;+\;
\ker\!\big(T_P^2 + \Tr_P\,T_P + \Nm_P\big)\ \Big] \ \subseteq\ S_2(\mathfrak n)\otimes\bbF_{11}.
\end{equation}
Nonvanishing of $V$ was the (flawed) criterion used in the retracted version; we now show it
is too weak.

\section{Refutation of the claimed match}\label{sec:refutation}

\paragraph{The criterion is too weak.} For each $P$, $\ker\big((T_P^2-\Tr_P T_P+\Nm_P)
(T_P^2+\Tr_P T_P+\Nm_P)\big)$ is the kernel of a degree-$4$ polynomial in $T_P$; over
$\bbF_{11}$ its dimension is roughly $\tfrac{4}{11}\dim S_2(\mathfrak n)$. Intersecting a few
such kernels can leave a nonzero $V$ for reasons unrelated to any eigenform whose twisted
eigenvalues match the fingerprint \eqref{eq:fingerprint}. One must test the eigenforms in $V$
directly.

\paragraph{The dim-$2$ space does not realize $\rho_1\oplus\rho_2$.} At
$\mathfrak n=\mathfrak p\cdot(2)$ one does find $\dim V=2$, but restricting $T_P$ to $V$ gives a
$2\times2$ matrix $C_P/\bbF_{11}$ (script \texttt{refine\_serre3.m}) for which, at most primes,
\[
\det C_P \ne \Nm_P,\qquad \operatorname{charpoly}(C_P)\notin\{q_P^+,q_P^-\},
\]
indeed $\operatorname{charpoly}(C_P)\in\{q_P^+,q_P^-\}$ at only $2/25$ test primes (both
degenerate, $\Nm_P=0$), and $C_P$ usually has a \emph{repeated} eigenvalue (e.g.\ $(y-7)^2$).
So $V$ is not semisimple and is not a sum of two genuine Hecke eigenforms realizing
$\rho_1,\rho_2$.

\paragraph{No eigenform matches, even up to twist.} The untwisted intersection
$V^+:=\bigcap_P\ker(T_P^2-\Tr_P T_P+\Nm_P)$ vanishes at level $\mathfrak p$ and at
$\mathfrak p\cdot(2)$ (scripts \texttt{refine\_serre4.m}, \texttt{refine\_serre5.m}). The unique
twist-robust survivor at level $\mathfrak p$ is a \emph{single} eigenform $f$ ($\dim V=1$); its
Hecke eigenvalue satisfies
\[
a_P \in \{\pm t_1,\pm t_2\}\quad\text{at all }120/120\text{ tested primes,}
\]
a genuine structural fact (probability $\approx(4/11)^{106}$ by chance). Writing
$\varepsilon(P)=+1$ if $a_P$ is a root of $q_P^+$ and $-1$ if of $q_P^-$, the $94$ unambiguous
signs are \emph{inconsistent with every quadratic ray-class character} of conductor supported
in $\{2^{\le6},3,5,11,66179,\infty\}$ (script \texttt{refine\_charfit.m}); since $K$ has narrow
class number $1$ there are no unramified quadratic characters either. So $\varepsilon$ is not a
quadratic Hecke character.

\begin{proposition}\label{prop:no-match}
No Hilbert modular eigenform of parallel weight $2$, trivial character, and level
$\mathfrak p\cdot(2)^a$ realizes $\rho_1$ or $\rho_2$.
\end{proposition}

\begin{proof}[Reason]
Such a form has $\det\rho_f=\chicyc=\det\rho_i$, so $\rho_f\cong\rho_i\otimes\chi$ forces
$\chi^2=1$: the only admissible twist is quadratic. A non-trivial nebentypus would change the
determinant and so cannot occur. But a quadratic twist relating $f$ to $\rho_i$ is exactly what
is excluded above (the sign $\varepsilon$ is not a quadratic character), and the untwisted case
is excluded by $V^+=0$.
\end{proof}

The residual relation $a_P^2\in\{t_1^2,t_2^2\}$ (the $120/120$ fact) is the twist-invariant,
projective ($\mathrm{Sym}^2$) statement, with the square jumping between $\rho_1$ and $\rho_2$;
it does \emph{not} imply that either $\rho_i$ is modular.

\begin{remark}
\textbf{Status.} The mod-$11$ Serre-modularity match for Curve~A is \emph{not} established.
The structural content (Propositions~\ref{prop:A},~\ref{prop:struct}) stands and is the
reusable output of this work. Natural next steps: a genuine newform decomposition with
eigensystems compared in the correct residue field (recall $11$ \emph{splits} in $K$, so a
matching newform would have $11$ split in its Hecke field and place both reductions in $V^+$,
which is empty at the levels examined); searching other levels; and interpreting the
$a_P^2$-relation. Reproducibility: \texttt{mod11\_rep.m}, \texttt{mod11\_trnm.txt} (target
fingerprint), and the refutation scripts \texttt{refine\_serre[3,4,5].m},
\texttt{refine\_charfit.m} with raw signs in \texttt{refine\_serre5\_signs.txt}.
\end{remark}

\section{The corrected modularity statement}\label{sec:derived}

Although we cannot exhibit the eigenform, the structural computations pin down what the correct
statement must be. Assuming Serre's conjecture over $K=\bbQ(\sqrt5)$ (each $\rho_i$ is
$2$-dimensional, irreducible, and odd, so it applies), each $\rho_i$ is modular, and its weight,
character and level are forced up to a single bounded constant.

\begin{theorem}[conditional on Serre's conjecture over $\bbQ(\sqrt5)$]\label{thm:derived}
Each of $\rho_1,\rho_2$ is modular by a Hilbert modular eigenform over $K$ of parallel weight
$(2,2)$, trivial nebentypus, and level $\mathfrak p\cdot(2)^{4}$, realized untwisted
($\det=\chicyc$). Here $a_2(\rho_i)=4$ is the conductor exponent at the inert prime $2$
(Section~\ref{sec:cond2}).
\end{theorem}

The parameters are forced as follows.
\begin{itemize}
\item\textbf{Determinant $\chicyc$:} the two Frobenius eigenvalues of each piece multiply to
$N(P)$ at all $132$ primes of type $[1,1,2]$ (det-clean).
\item\textbf{Trivial nebentypus:} $\det\rho_f=\psi\,\chicyc^{k_0-1}$, so weight $2$ with
$\det=\chicyc$ forces $\psi$ trivial. Nebentypus $\chicyc$ would give $\det=\chicyc^{2}$; as
$\chicyc$ has order $10$ it has no square root among its powers, so no twist of $\rho_i$ has that
determinant.
\item\textbf{Weight $(2,2)$:} $A$ is \emph{ordinary} at both primes above $11$ ($L_{\mathfrak
l}\bmod 11$ has degree $2$, i.e.\ two unit roots). Ordinary $+$ good reduction $\Rightarrow$
peu-ramifi\'ee; a subquotient of a finite-flat group scheme is finite-flat (as $e=1<11-1$, by
Raynaud); and $\det=\chicyc$ forces Hodge--Tate weights $\{0,1\}$. Hence Serre weight $2$ at both
primes.
\item\textbf{Level $\mathfrak p\cdot(2)^{a}$:} pure $2$-power levels $(2)^{a}$ have empty match,
so $\rho_i$ is ramified at $\mathfrak p$; and the untwisted match is empty at $\mathfrak
p\cdot(2)^{0,1,2}$, so $a\ge 3$.
\end{itemize}

\begin{remark}
Since the untwisted match is empty at $2$-part $\le 2$ for both pieces, $a_2(\rho_1),a_2(\rho_2)
\ge 3$, whence (Artin conductor is additive) $a_2(A)=a_2(\rho_1)+a_2(\rho_2)\ge 6$. Wild inertia
at $2$ acts through a $2$-group of order $\le 16$ (the $2$-Sylow of $\mathrm{GL}_2(\bbF_{11})$).
\end{remark}

\section{The conductor at $2$}\label{sec:cond2}

We determine $c:=a_{\mathfrak p_2}(A)$ numerically from the functional equation. Magma provides
neither regular models nor minimisation for genus $2$ over number fields, so we work with the
degree-$8$ $L$-function $L(A/K,s)=L(W/\bbQ,s)$, $W=\mathrm{Res}_{K/\bbQ}A$, whose analytic
conductor is
\[
N(W)=|d_K|^{4}\cdot N_{K/\bbQ}(\mathfrak p\cdot(2)^{c})=5^{4}\cdot 66179\cdot 4^{c}.
\]

\paragraph{The right twist.} The reconstruction \texttt{Genus2CurveFromIgusa} returns a curve
$A_{\mathrm{rec}}$ that is \emph{additive of potentially good reduction} at
$499,13711,88301,231481,\dots$ (a spurious quadratic twist by the residual model content $g$;
the Igusa criterion misses these because potential-good passes it). The correctly twisted curve
$A=A_{\mathrm{rec}}\otimes\chi_g$, with $\chi_g$ the quadratic character of $K(\sqrt g)$, is good
at all those primes, so its conductor is exactly $\mathfrak p\cdot(2)^{c}$ and $L(A/K,s)$ is
tractable. Its Euler factors are the $\chi_g$-twist of those of $A_{\mathrm{rec}}$:
$L^{A}_P(T)=L^{A_{\mathrm{rec}}}_P(\chi_g(P)\,T)$.

\paragraph{Result.} Building the Euler product to norm $3\cdot10^{5}$ ($25\,981$ factors) and
running \texttt{CheckFunctionalEquation} at precision $3$ (\texttt{cfe\_final2.m}):
\[
\begin{array}{c|cccc}
c & 6 & 7 & \mathbf{8} & 9\\\hline
|\mathrm{CFE}| & 0.10 & 0.29 & \mathbf{0.0020} & 0.23
\end{array}
\]
$c=8$ closes the functional equation about two orders of magnitude more sharply than any
neighbour, and is consistent with $a_2(A)\ge 6$. Hence $a_{\mathfrak p_2}(A)=8$, and
(symmetrically) $a_2(\rho_1)=a_2(\rho_2)=4$, giving the level $\mathfrak p\cdot(2)^4$ of
Theorem~\ref{thm:derived}. This is a numerical (functional-equation) determination, not a
rigorous proof.

\section{An explicit form for $\rho_1$}\label{sec:explicit}

The level-$\mathfrak p$ eigenform with $a_P^2\in\{t_1^2,t_2^2\}$, dismissed above as coincidental,
is in fact a genuine twist of $\rho_1$; it was missed only because the character search there
omitted the primes $499,13711,88301,231481$ where the reconstruction's spurious twist $\chi_g$
ramifies. Sweeping the quadratic characters $\chi_d$ ramified only at $2$ and matching
$A\otimes\chi_d$ by the untwisted filter, exactly one survives --- $\chi_{-1}$, the character of
$K(i)/K$ --- and already at level $\mathfrak p$ (verified over $80$ primes).

\begin{proposition}\label{prop:g}
$\rho_1\cong \rho_{g}\otimes\chi$, where $\chi=\chi_g\cdot\chi_{-1}$ is the quadratic character of
$K(\sqrt{-g})$ and $g$ is an explicit Hilbert modular newform over $K$ of weight $(2,2)$, trivial
nebentypus, and level $\mathfrak p$ (norm $66179$): it is one of the four newform orbits of
$S_2(\mathfrak p)$, with Hecke field $E=\bbQ[x]/(x^5+8x^4+19x^3+4x^2-32x-23)$, in which $11$ is
totally ramified.
\end{proposition}

Since $11$ is totally ramified in $E$, $g$ has a single mod-$11$ reduction (residue field
$\bbF_{11}$), which is $\rho_1\otimes\chi$; consistently, $a_2(\rho_1)=4=2\cdot(\text{conductor of
}\chi_{-1}\text{ at }2)$, so the twist $\chi_{-1}$ undoes the wild $2$-ramification and lowers the
level from $\mathfrak p\cdot(2)^4$ to $\mathfrak p$.

\begin{remark}
The two pieces behave differently at $2$: for $\rho_2$ no quadratic twist ramified at $2$ reaches
level $\mathfrak p\cdot(2)^2$, i.e.\ $a_2(\rho_2\otimes\chi_d)\ge 3$ for all such $\chi_d$, so
$\rho_2$ is twist-minimal (supercuspidal-like) at $2$. Its form lives at level $\mathfrak
p\cdot(2)^3$ (dimension $\approx 88000$), where a single Hecke operator already exceeds available
memory; $\rho_2$ is thus characterised but not yet exhibited.
\end{remark}

\end{document}
